Fast Sorting on Modern Computers

### 2005-08-02 docs/tech_report/method.tex
fixed up the references
wrote about testing

%TODO this is going to have to be updated
\begin{figure}
\scalebox{0.57}{\includegraphics*[0pt,165pt][600pt,802pt]{flowchart}}
\caption{Flowchart describing the development of an algorithm}
\label{Flowchart}
\end{figure}

@@

%TESTING - what version
\subsection{Valgrind}
\label{Valgrind}
\cc{Valgrind} is a set of tools which are used for debugging and profiling.  By
default it works as a memory checker, and detects errors such as the use of
uninitialised memory and accessing memory off the end of an array. This use
makes it particularly useful for bounds checking. No configuration was required
for this, and simply running \cc{valgrind a.out} ran the program through
valgrind. The homepage for valgrind is at \n{http://valgrind.kde.org/}.

%TESTING - what version
\subsection{PapiEx}
\cc{PapiEx} is a tool used to access performance counters on a Pentium 4. It
uses the PAPI framework to count certain characteristics across a program's
execution. We used it to count the number of cycles a sort took to run. This
provided an accurate measure of how fast a sort executes on a Pentium 4. The
PapiEx homepage is at \n{http://icl.cs.utk.edu/~mucci/papiex/}.

@@

>>>>

%TODO this is going to have to be updated - or is it
\begin{figure}
\scalebox{0.57}{\includegraphics*[0pt,165pt][600pt,802pt]{flowchart}}
\caption{Flowchart describing the development of an algorithm}
\label{Flowchart}
\end{figure}

@@

\subsection{Valgrind}
\label{Valgrind}
\cc{Valgrind} is a set of tools which are used for debugging and profiling.  By
default it works as a memory checker, and detects errors such as the use of
uninitialised memory and accessing memory off the end of an array. This use
makes it particularly useful for bounds checking. No configuration was required
for this, and simply running \cc{valgrind a.out} ran the program through
valgrind. The homepage for valgrind is at \n{http://valgrind.kde.org/}.

\subsection{PapiEx}
\cc{PapiEx} is a tool used to access performance counters on a Pentium 4. It
uses the PAPI framework to count certain characteristics across a program's
execution. We used it to count the number of cycles a sort took to run. This
provided an accurate measure of how fast a sort executes on a Pentium 4. The
PapiEx homepage is at \n{http://icl.cs.utk.edu/\textasciitilde{}mucci/papiex/}.

@@

Our tests were designed so that we could compare our results to LaMarca's, and
so our results true in the general case. Both our testing method and our data
sets were chosen with these goals in mind.

\subsection{SimpleScalar Simulations}

We tested each of our sorts using SimpleScalar, using set sizes from 4096 to
4194304, increasing by a factor of two each time. Each of these set sizes was
simulated using \cc{sim-cache} and \cc{sim-bpred} in a variety of
configurations.

The SimpleScalar settings were chosen to be similar to LaMarca's simulations.
LaMarca used the \n{Atom} simulator, which was used to test software for the DEC
AlphaStation. It simulated a 2MB direct-mapped cache with a 32 byte cache line.
Other settings were also chosen to correspond to the DEC Alpha: 8KB level 1 data
cache, 8KB level 1 instruction cache and shared instruction and data level 2
cache.

In order to be able to extrapolate more information, we increased the number of
cache simulations. We simulated both 2MB and 4MB caches, configured as both
direct-mapped and fully-associative.

We chose the a similar approach for our branch predictor caches. We aimed to
simulate bimodal predictors as well as 2 level adaptive, in order to be able to
compare them together. We simulated bimodal predictors with a 512, 1024 and
2048 byte table, and 2 level adaptive predictors with 512, 1024, 2048 and 4096
entries.

\subsection{Data}

Both our SimpleScalar simulations and our software predictor tests used truly
random data from \n{http://www.random.org}. The software predictor used ten
sections containing 4194304 \cc{unsigned integer}s, and the SimpleScalar
simulations used the same ten sections. When SimpleScalar simulations were run
on sized smaller than 4194304 keys, only the left hand side of the sections were
used.

\subsection{PaPiex}

Our papiex tests ran significantly faster than either of our other simulations,
and so we were able to run them over a significantly greater quantity of data.
Each sort, and each size from 4096 to 4194304 key, was run 1024 times, using the
system's random number generator to generate data, using seeds of between 0 and
1023.

%TODO did i say 128 byte elsewhere?
These simulations were run a on Pentium 4 1.8Ghz, which had an 8-way, 256K level
2 cache, with a 64-byte cache line. The level 1 caches are seperate 8K 4-way
associative caches.

The instruction pipeline of the Pentium 4 is 20 stages long, and so the branch
misprediction penalty of the Pentium 4 approaches 20 cycles. The Pentium 4's
static branch prediction strategy is backward taken and forward not taken; its
dynamic strategy is unspecified: though \cite{Intel248966-010} mentions a branch
history so it is likely a two-level adaptive predictor.

\subsection{32-bit Integers}
LaMarca used 64-bit \cc{integer}s, as his test machine was a 64-bit system.
Since our machine is 32-bit, and SimpleScalar uses 32-bit addressing on our
architecture, LaMarca's results will not be identical to ours. Rather, the
number of cache misses in his results should be halved to be compared to our;
when this is not the case, it is indicated.

\subsection{Filling Arrays}

It takes time to fill arrays, and this can distort the relationship between the
results for small sizes, and larger sizes, for which this time is amortized. As
a result, we measured this time, and subtracted it from our results. None of the
graphs show this time.

However, this practice also distorts the results. It removes compulsory cache
misses from the results, so long as the data fits in the cache. When the data
does not fit in the cache, the keys from the start of the array are faulted,
with the effect that capacity misses reoccur, which are not discounted. As a
result, on the left hand side of cache graphs, a number of compulsory misses are
removed, though they are not on the right hand side. Similarly, LaMarca does not
do this, and so this needs to be taken into account when comparing his results
to ours.
